Short-time simulations of open collective quantum spin systems

### 2019-04-12 soc_squeezing/collective_paper/collective_spin.tex
added abstract, reference, and changed documentclass to revtex4-2

\documentclass[pra,reprint,longbibliography]{revtex4-1}

% linking references
\usepackage{hyperref}
\hypersetup{
  breaklinks=true,
  colorlinks=true,
  linkcolor=blue,
  filecolor=magenta,
  urlcolor=cyan,
}

@@

% for strikeout text
% normalem included to prevent underlining titles in the bibliography
\usepackage[normalem]{ulem}

% for leaving notes in the text
\newcommand{\note}[1]{\textcolor{red}{#1}}

@@

\begin{abstract}
  Abstract goes here.
\end{abstract}

@@

\begin{figure}
  \centering
  \subfloat[Squeezing under coherent dynamics
  \label{fig:squeezing_coherent}]
  {\includegraphics{coherent.pdf}} \\
  \subfloat[Squeezing with decoherence:
  $\gamma_-=\gamma_+=\gamma_\z=\chi$
  \label{fig:squeezing_incoherent}]
  {\includegraphics{decoherence_weak.pdf}}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$
    under \sref{fig:squeezing_coherent} coherent and
    \sref{fig:squeezing_incoherent} incoherent evolution, computed
    using exact methods (solid lines), quantum trajectory simulations
    (dots), and the TST expansion in Eq.~\eqref{eq:order_truncation}
    with $M=35$ (darker dashed lines).  Solid circles mark the times
    at which the TST expansion gives an unphysical result with
    $\xi^2<0$.}
  \label{fig:benchmarking}
\end{figure}

@@

The main lesson from Figure \ref{fig:benchmarking} is that the TST
expansion gives essentially exact results right up until a sudden and
drastic departure that is simple to diagnose by inspection.  This
drastic departure occurs because the TST expansion neglects
high-weight (i.e.~large-$\abs{\v m}$) operators $\S_{\v m}$ whose
contributions to a Heisenberg operator $\O\p{t}$ of interest
eventually become non-negligible.  For sufficiently large times $t$,
these individual contributions may generally be large compared to the
value of $\bk{\O\p{t}}$, in which case their truncation will yield
non-physical results.  This breakdown mechanism is also the reason for
prolonged agreement between the TST expansion and benchmarking methods
when squeezing the initial state $\ket{-\Z}$ rather than $\ket{\X}$:
for initial states $\ket{-\Z}$, all initial-time correlators
$\bk{\S_{\v m}}_{t=0}$ vanish unless $m_+=m_-=0$ (see Appendix
\ref{sec:initial_conditions}), so there is a substantially smaller
number of neglected terms with non-zero contribution to the correlator
$\bk{\O\p{t}}$.

@@

As a final example of collective-spin physics that is numerically
accessible via the TST expansion of Heisenberg operators, we consider
the calculation of collective-spin two-time correlation functions.  In
particular, we consider the effect of decoherence on the short-time
behavior of the two-time connected correlator
\begin{align}
  C\p{t}
  \equiv \f1S\sp{\bk{S_+\p{t}S_-\p{0}}-\bk{S_+\p{t}}\bk{S_-\p{0}}}
  \label{eq:two_time}
\end{align}
in the context of the squeezing models in Section \ref{sec:squeezing}.
In an equilibrium setting, correlation functions of this sort contain
information about the linear response of Heisenberg operators to
perturbations of a system[CITE]\cite{CITE}; in a non-equilibrium
setting, they contribute to short-time linear response (see Appendix
\ref{sec:linear_response}).  Similar correlators have made appearances
as order parameters that diagnose time-crystalline phases of
matter\cite{tucker2018shattered}, and related multi-time correlators,
known as out-of-time-ordered correlators, are commonly examined for
signatures of quantum chaos and information
scrambling[CITE]\cite{CITE}.

\begin{figure*}
  \centering
  \includegraphics{Sp_Sm.pdf}
  \caption{Real (solid lines) and imaginary (dashed lines) parts of
    the two-time connected correlator $C\p{t}$ defined in
    Eq.~\eqref{eq:two_time} for $N=10^4$ spins initially in the
    polarized state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
    evolving under the squeezing Hamiltonians in
    Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.  Results are shown for time
    evolution without decoherence (leftmost panel), as well as with
    decoherence at rates $\gamma_n$ or $\Gamma_n$, respectively
    corresponding to single-spin or collective jump operators
    $s_n^{(j)}$ or $S_n$.  All axes in the sub-figures are shared.  At
    short times, the correlations captured by $C\p{t}$ are suppressed
    by single-spin decay ($\gamma_-$), collective decay ($\Gamma_-$),
    and single-spin dephasing ($\gamma_\z$); these correlations are
    relatively insensitive to collective dephasing ($\Gamma_\z$), and
    can change sign or grow from spontaneous single-spin ($\gamma_+$)
    or collective ($\Gamma_+$) excitation.}
  \label{fig:two_time}
\end{figure*}

Figure \ref{fig:two_time} shows the behavior of $C\p{t}$ for $N=10^4$
spins, initially in the state
$\ket\X\propto\p{\ket\up+\ket\dn}^{\otimes N}$, evolving under the
squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} both
with and without various sources of decoherence.  The two-time
correlator $C\p{t}$ degrades at short times under certain types of
decoherence, but seems relatively insensitive to collective dephasing
(with jump operator $S_\z$), and can qualitatively change behavior or
grow in the presence of single-spin or collective excitation (with
jump operators $s_+^{(j)}$ or $S_+$).  Figure \ref{fig:two_time}
serves as an example for the type of quantity and behavior that is
accessible at short times with the TST expansion.  This example is
straightforward to extend to multi-time and out-of-time-ordered
correlators (OTOCs), as well as equilibrium settings.

@@

We have presented a new method for computing correlators in collective
spin systems at short times.  This method is based on truncating a
short-time expansion of Heisenberg operators, and can access
correlators on time scales that are relevant to metrological
applications such as spin squeezing.  In order to evaluate the
truncated short-time (TST) expansion of Heisenberg operators, we have
computed the structure constants of a collective spin operator
algebra, which we hope will empower future analytical and numerical
studies of collective spin systems.  Even though we considered only
non-equilibrium spin-squeezing processes in this work, our method can
be applied directly in an equilibrium setting, and is straightforward
to generalize to systems such as trapped ions and optical cavities
with collective spin-boson interactions.  In such contexts, our method
may be used to benchmark the short-time effects of decoherence, or
compute quantities such as out-of-time-ordered correlators (OTOCs)
that diagnose the onset of quantum chaos and information scrambling.

\section{Acknowledgments}

We acknowledge helpful discussions with Robert Lewis-Swan and Kris
Tucker.  This work is supported by the Air Force Office of Scientific
Research (AFOSR) grant FA9550-18-1-0319; the AFOSR Multidisciplinary
University Research Initiative (MURI) grant; the Defense Advanced
Research Projects Agency (DARPA) and Army Research Office (ARO) grant
W911NF-16-1-0576; the National Science Foundation (NSF) grant
PHY-1820885; JILA-NSF grant PFC-173400; and the National Institute of
Standards and Technology (NIST).

\newpage
\onecolumngrid
\appendix

@@

  retrospect, it would likely have been nicer to have all
  $s_\mu^{(j)}$ operators on the right throughout these calculations,
  due to the enhanced symmetry expressions would have with respect to
  Hermitian conjugation.  In any case, we provide the final result of
  this section in both ordering conventions, and therefore feel no
  need to reproduce these calculations with a different ordering of
  spin operators.}.  This choice of ordering compels us to expand

>>>>

\documentclass[pra,twocolumn,longbibliography]{revtex4-2}

% linking references
\usepackage{hyperref}
\hypersetup{
  breaklinks=true,
  colorlinks=true,
  linkcolor=blue,
  urlcolor=cyan,
}

@@

@@

\begin{abstract}
  We present a new method to compute short-time expectation values in
  collective spin systems with permutationally-symmetric decoherence.
  Our method is based on a Taylor expansion of a formal solution to
  the equations of motion for Heisenberg operators.  This expansion
  can be truncated at finite order to obtain nearly exact results at
  short times that are relevant for metrological applications such as
  spin squeezing.  In order to evaluate the expansion for Heisenberg
  operators, we compute the structure constants of a collective SU(2)
  operator algebra.  We demonstrate the utility of our method by
  computing spin squeezing and two-time correlation functions in
  strong-decoherence regimes that are inaccessible to existing
  numerical methods.  Our method is straightforward to generalize to
  the case of a collective spin coupled to bosonic modes, as can occur
  in trapped ion and optical cavity experiments, and may be used to
  compute quantities such as out-of-time-ordered correlators that
  contain short-time signatures of quantum chaos and information
  scrambling.
\end{abstract}

@@

\begin{figure}
  \centering
  \subfloat[Squeezing with coherent dynamics
  \label{fig:squeezing_coherent}]
  {\includegraphics{coherent.pdf}} \\
  \subfloat[Squeezing with decoherence:
  $\gamma_-=\gamma_+=\gamma_\z=\chi$
  \label{fig:squeezing_incoherent}]
  {\includegraphics{decoherence_weak.pdf}}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$
    under \sref{fig:squeezing_coherent} coherent and
    \sref{fig:squeezing_incoherent} incoherent evolution, computed
    using exact methods (solid lines), quantum trajectory simulations
    (dots), and the TST expansion in Eq.~\eqref{eq:order_truncation}
    with $M=35$ (darker dashed lines).  Solid circles mark the times
    at which the TST expansion gives an unphysical result with
    $\xi^2<0$.}
  \label{fig:benchmarking}
\end{figure}

@@

The main lesson from Figure \ref{fig:benchmarking} is that the TST
expansion gives essentially exact results right up until a sudden and
drastic departure that can be diagnosed by inspection.  This drastic
departure occurs because the TST expansion neglects high-weight
(i.e.~large-$\abs{\v m}$) operators $\S_{\v m}$ whose contributions to
a Heisenberg operator $\O\p{t}$ of interest eventually become
non-negligible.  For sufficiently large times $t$, these individual
contributions may generally be large compared to the value of
$\bk{\O\p{t}}$, in which case their truncation will yield non-physical
results.  This breakdown mechanism is also the reason for prolonged
agreement between the TST expansion and benchmarking methods when
squeezing the initial state $\ket{-\Z}$ rather than $\ket{\X}$: for
initial states $\ket{-\Z}$, all initial-time correlators
$\bk{\S_{\v m}}_{t=0}$ vanish unless $m_+=m_-=0$ (see Appendix
\ref{sec:initial_conditions}), so there is a substantially smaller
number of neglected terms with non-zero contribution to the correlator
$\bk{\O\p{t}}$.

@@

\begin{figure*}
  \centering
  \includegraphics{two_time.pdf}
  \caption{Real (solid lines) and imaginary (dashed lines) parts of
    the two-time connected correlator $C\p{t}$ defined in
    Eq.~\eqref{eq:two_time} for $N=10^4$ spins initially in the
    polarized state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
    evolving under the squeezing Hamiltonians in
    Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.  Results are shown for time
    evolution without decoherence (leftmost panel), as well as with
    decoherence at rates $\gamma_n$ or $\Gamma_n$, respectively
    corresponding to single-spin or collective jump operators
    $s_n^{(j)}$ or $S_n$.  All axes in the sub-figures are shared.  At
    short times, the correlations captured by $C\p{t}$ are suppressed
    by single-spin decay ($\gamma_-$), collective decay ($\Gamma_-$),
    and single-spin dephasing ($\gamma_\z$); these correlations are
    relatively insensitive to collective dephasing ($\Gamma_\z$), and
    can change sign or grow from spontaneous single-spin ($\gamma_+$)
    or collective ($\Gamma_+$) excitation.}
  \label{fig:two_time}
\end{figure*}

As a final example of collective-spin physics that is numerically
accessible via the TST expansion of Heisenberg operators, we consider
the calculation of collective-spin two-time correlation functions.  In
particular, we consider the effect of decoherence on the short-time
behavior of the two-time connected correlator
\begin{align}
  C\p{t}
  \equiv \f1S\sp{\bk{S_+\p{t}S_-\p{0}}-\bk{S_+\p{t}}\bk{S_-\p{0}}}
  \label{eq:two_time}
\end{align}
in the context of the squeezing models in Section \ref{sec:squeezing}.
In an equilibrium setting, correlation functions of this sort contain
information about the linear response of Heisenberg operators to
perturbations of a system; in a non-equilibrium setting, they
contribute to short-time linear response (see Appendix
\ref{sec:linear_response}).  Similar correlators have made appearances
as order parameters that diagnose time-crystalline phases of
matter\cite{tucker2018shattered}, and related multi-time correlators,
known as out-of-time-ordered correlators, are commonly examined for
signatures of quantum chaos and information
scrambling\cite{maldacena2016bound}.

Figure \ref{fig:two_time} shows the behavior of $C\p{t}$ for $N=10^4$
spins, initially in the state
$\ket\X\propto\p{\ket\up+\ket\dn}^{\otimes N}$, evolving under the
squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} both
with and without various sources of decoherence.  The two-time
correlator $C\p{t}$ degrades at short times under certain types of
decoherence, but seems relatively insensitive to collective dephasing
(with jump operator $S_\z$), and can qualitatively change behavior or
grow in the presence of single-spin or collective excitation (with
jump operators $s_+^{(j)}$ or $S_+$).  Figure \ref{fig:two_time}
serves as an example for the type of quantity and behavior that is
accessible at short times with the TST expansion.  This example is
straightforward to extend to multi-time and out-of-time-ordered
correlators, as well as equilibrium settings.

@@

We have presented a new method for computing correlators in collective
spin systems at short times.  This method is based on truncating a
short-time expansion of Heisenberg operators, and can access
correlators on time scales that are relevant to metrological
applications such as spin squeezing.  In order to evaluate the
truncated short-time (TST) expansion of Heisenberg operators, we have
computed the structure constants of a collective spin operator
algebra, which we hope will empower future analytical and numerical
studies of collective spin systems.  Even though we considered only
non-equilibrium spin-squeezing processes in this work, our method can
be applied directly in an equilibrium setting, and is straightforward
to generalize to systems such as trapped ions and optical cavities
with collective spin-boson interactions.  In such contexts, our method
may be used to benchmark the short-time effects of decoherence, or
compute quantities such as out-of-time-ordered correlators that
diagnose the onset of quantum chaos and information scrambling.

\begin{acknowledgments}
  We acknowledge helpful discussions with Robert Lewis-Swan and Kris
  Tucker.  This work is supported by the Air Force Office of
  Scientific Research (AFOSR) grant FA9550-18-1-0319; the AFOSR
  Multidisciplinary University Research Initiative (MURI) grant; the
  Defense Advanced Research Projects Agency (DARPA) and Army Research
  Office (ARO) grant W911NF-16-1-0576; the National Science Foundation
  (NSF) grant PHY-1820885; JILA-NSF grant PFC-173400; and the National
  Institute of Standards and Technology (NIST).
\end{acknowledgments}

\onecolumngrid
\appendix

@@

  retrospect, it may have been nicer to push all $s_\mu^{(j)}$
  operators to the right throughout these calculations, due to the
  enhanced symmetry expressions would have with respect to Hermitian
  conjugation.  In any case, we provide the final result of this
  section in both ordering conventions, and therefore feel no need to
  reproduce these calculations with a different ordering of spin
  operators.}.  This choice of ordering compels us to expand
